\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-01-12 paper.qmd

```{r}
#| label: fig-rm-lg
#| fig-cap: "Distribution of edge lengths and wireframes in \\gD{} and $7\\text{-}D$ with Delaunay triangulation, and two choices of long edge removal: (b) benchmark = 2.5$a_1$ and (c) benchmark = 2$a_1$, where $a_1$ = 0.09."
#| fig-height: 15
#| fig-width: 15

@@

<!-- DHC: Will need to have more bins in each of the methods to illustrate this properly. -->

To illustrate a difference in how the methods organise points in the NLDR layout, simulated $4\text{-}D$ data having five Gaussian clusters is used. @fig-five-gau-projs a1, b1, c1 show the \gD{} layouts for (a) tSNE, (b) UMAP, and (c) PaCMAP, respectively. The default hyper-parameters are used. All three methods show the five clusters, with varying degrees of separation.

The models are fitted to these layouts, but we focus on a single cluster to illustrate the curious detail. @fig-five-gau-projs a2, b2, c2 show the fitted models in a projection of the $4\text{-}D$ space. In this projection the difference between methods can be seen. These clusters are fully $4\text{-}D$ in nature, so we would expect the model to be a *crumpled \gD{} sheet* that stretches in all four dimensions. This is what is observed for tSNE and UMAP. The curious detail is that the model for PaCMAP is closer to a *pancake* in shape! What this means is that there has to be some ordering of points in the \gD{} PaCMAP layout that induces the flat model, likely that the points are organised by the global principal components. So the layout of the model in $4\text{-}D$ doesn't reflect the dimension of each cluster. DHC: is there something in the documentation that might help to explain what has happened?

This pattern can also be observed by examining the error from the model fit (@fig-five-gau-projs a3, b3, c3). With tSNE and UMAP the error is relatively uniformly distributed, but with PaCMAP there is more error in the center of the \gD{} layout, reflecting that there are more points in the middle that are further from the fitted model.

@@

```{r}
#| label: fig-one-dens_clust-error
#| fig-height: 10
#| fig-width: 10
#| fig-pos: H
#| fig-cap: "The sparse end of the structure exhibits a high model error, while the dense end shows a low model error. The tSNE layouts of the C-shaped cluster, with uniform points (a1) and dense points (b1) are colored according to their high-dimensional error, along with their selected projections (a2 and b2), are displayed. Darker colors represent a high error, while lighter colors indicate a low error. The uniformly distributed points demonstrate random model errors throughout the structure, whereas the dense structure shows large model error at the sparse end."

@@

### Single-cell gene expression
<!--
- NLDR view used to illustrate clusters
- Use our method to assess is it a reasonable representation
- Demonstrate that it is not
- Illustrate how to use our method to get a better representation
-->

In the field of single-cell studies, a common analytical task involves clustering to identify groups of cells with similar expression profiles. NLDR methods are commonly used to display clusters, and help to verify the results. For example, @chen2023 illustrates the use of UMAP to identify clusters in Human Peripheral Blood Mononuclear Cells (PBMC3k). There are $2622$ single cells. First 9 principal components are used to generate the UMAP. @fig-NLDR-variety (a) is the reproduction of the published plot. The objective is to assess the published layout, and if it does not accurately represent the three clusters with small separations of the PBMC3k dataset (@fig-model-pbmc-author-proj (a2)), then select a reasonable \gD{} layout.

@@

As a result, it is necessary to find an appropriate layout for the dataset. MSE for different binwidths ($a_1$) using tSNE, UMAP, PHATE, PaCMAP, and TriMAP with various (hyper-)parameter settings were computed (@fig-pbmc-mse). Layouts c, d, and e, which show small separations between clusters, are universally optimal. However, layout d performs well with smaller binwidths and poorly with larger binwidths. On the other hand, layout e performs well with larger binwidths. Layout c was selected for further analysis due to its better (hyper-)parameter selection for the same method of the published plot.

@@

```{r}
#| echo: false
#| fig-cap: "Model in \\gD{}, on the layout a and b, and two views of the fit in projections from $9\\text{-}D$, for the PBMC3k data ($n =  2622$ and $p = 9$). Layout a shows three-well separated clusters with big separation, while layout b shows close clusters. In $9\\text{-}D$, the data shows three clusters with two of them being close and one separated from the others. This is why b1, b2, and b3 shows a connected edge between the close clusters, as proven by @fig-pbmc-mse as a reasonable layout. Viewing the fitted model in $9\\text{-}D$, helps to see some unobserved patterns of the data: (i) dense points, and (ii) non-linear clusters. Videos of the langevitour animations are available at <https://youtu.be/0cKX_HG_n0k> and <https://youtu.be/KhJvsRtaX04> respectively."
#| label: fig-model-pbmc-author-proj
#| fig-pos: H
#| fig-width: 10
#| fig-height: 15
#| out-width: 80%

@@

In the field of single-cell studies, clustering is a common analytical task used to identify groups of cells with similar expression profiles. Non-linear dimensional reduction (NLDR) methods are frequently employed to visualize these clusters and help validate the results. However, it is well known that the 2D embeddings produced by t-SNE and UMAP may not accurately reflect the similarities among cell clusters.

To address this challenge, @xia2023 introduces a statistical method called scDEED, which detects unreliable cell embeddings generated by \gD{} embedding techniques. scDEED calculates a reliability score for each cell embedding based on the similarity between the cell’s \gD{} embedding neighbors and its neighbors prior to embedding. By identifying embeddings with low reliability scores as dubious and those with high scores as trustworthy, scDEED minimizes the number of questionable cell embeddings. This approach also provides clear guidance for optimizing the hyperparameters of an embedding method.

To verify that the \gD{} cell embeddings became better aligned with the cell types after scDEED’s optimization, the Human Peripheral Blood Mononuclear Cells (PBMC) dataset was used. It contained $31,021$ cells with cell type labels, and the gene expression levels were in the unit of log-transformed UMI count per $10,000$. They focused on three sequencing methods (inDrops, DropSeq, and SeqWell) and four common cell types Cytotoxic T cell, CD4+T cell, CD14+ Monocyte, and B cell.

@@

@@

```{r}
#| label: fig-pbmc-mse-umap
#| fig-pos: H
#| out-width: 100%
#| fig-cap: "Assessing which of the 2 UMAP layouts with different hyperparameter settings (n_neighbors: $30$, min_dist: $0.3$ (red), n_neighbors: $80$, min_dist: $0.5$ (blue)) on the PBMC data is the better representation using MSE for varying binwidth ($a_1$). Colour  used for the lines and points in the left plot and in the scatterplots represents UMAP layout (a, b). Layout b is universally suboptimal. Layout b is the best choice."

@@

```{r}
#| echo: false
#| fig-cap: "Model in \\gD{}, on the layout a, and two views of the fit in projections from $50\\text{-}D$, for the PBMC data ($n =  5858$ and $p = 50$). Layout shows close clusters. In $50\\text{-}D$, the data shows "
#| label: fig-model-pbmc-author-proj-scdeed
#| fig-pos: H
#| fig-width: 15
#| fig-height: 5
#| out-width: 80%

>>>>

```{r}
#| label: fig-rm-lg
#| fig-cap: "XXX Shouldn't the middle one have an edge connecting inside a C, so that it is clearly not as good as c? XXX Distribution of edge lengths and wireframes in \\gD{} and $7\\text{-}D$ with Delaunay triangulation, and two choices of long edge removal: (b) benchmark = 2.5$a_1$ and (c) benchmark = 2$a_1$, where $a_1$ = 0.09."
#| fig-height: 15
#| fig-width: 15

@@

DHC: Will need to have more bins in each of the methods to illustrate this properly.

To illustrate a difference in how the methods organise points in the NLDR layout, simulated 4D data having five Gaussian clusters is used. @fig-five-gau-projs a1, b1, c1 show the 2D layouts for (a) t-SNE, (b) UMAP, and (c) PaCMAP, respectively. The default hyper-parameters are used. All three methods show the five clusters, with varying degrees of separation.

The models are fitted to these layouts, but we focus on a single cluster to illustrate the curious detail. @fig-five-gau-projs a2, b2, c2 show the fitted models in a projection of the 4D space. In this projection the difference between methods can be seen. These clusters are fully 4D in nature, so we would expect the model to be a *crumpled 2D sheet* that stretches in all four dimensions. This is what is observed for t-SNE and UMAP. The curious detail is that the model for PaCMAP is closer to a *pancake* in shape! What this means is that there has to be some ordering of points in the 2D PaCMAP layout that induces the flat model, likely that the points are organised by the global principal components. So the layout of the model in 4D doesn't reflect the dimension of each cluster. DHC: is there something in the documentation that might help to explain what has happened?

This pattern can also be observed by examining the error from the model fit (@fig-five-gau-projs a3, b3, c3). With t-SNE and UMAP the error is relatively uniformly distributed, but with PaCMAP there is more error in the center of the 2D layout, reflecting that there are more points in the middle that are further from the fitted model.

@@

```{r}
#| label: fig-one-dens_clust-error
#| fig-height: 10
#| fig-width: 10
#| fig-pos: H
#| fig-cap: "The sparse end of the structure exhibits a high model error, while the dense end shows a low model error. The tSNE layouts of the C-shaped cluster, with uniform points (a1) and dense points (b1), along with their selected projections (a2 and b2), are displayed. Darker colors represent a high error, while lighter colors indicate a low error. The uniformly distributed points demonstrate random model errors throughout the structure, whereas the dense structure shows large model error at the sparse end."

@@

XXX WHY THESE EXAMPLES. NEEDS AN OVERVIEW PARAGRAPH

### Single-cell gene expression {#sec-pbmc}
<!--
- NLDR view used to illustrate clusters
- Use our method to assess is it a reasonable representation
- Demonstrate that it is not
- Illustrate how to use our method to get a better representation
-->

XXX What is single cell data????

Clustering of single-cell data is used to identify groups of cells with similar expression profiles. NLDR often used to summarise the discovered clusters, and help to understand the results. The purpose of this example is to *illustrate how to use our method to help decide on an appropriate NLDR layout that accurately represents the data*. The cluster results of Human Peripheral Blood Mononuclear Cells (PBMC3k) in @chen2023 are examined. There are $2622$ single cells, with XXX gene expressions. Following their pre-processing, UMAP was performed using nine principal components. @fig-NLDR-variety (a) is the reproduction of the published plot, with colour indicating the clustering results overlaid on the NLDR layout. The question is whether this accurately represents the cluster structure in the data. Our method can help here, and also help to provide a better \gD{} layout, as needed.

@@

As a result, it is necessary to find an appropriate layout for the dataset. MSE for different binwidths ($a_1$) using tSNE, UMAP, PHATE, PaCMAP, and TriMAP with various (hyper-)~parameter settings were computed (@fig-pbmc-mse). Layouts c, d, and e, which show small separations between clusters, are universally optimal. However, layout d performs well with smaller binwidths and poorly with larger binwidths. On the other hand, layout e performs well with larger binwidths. Layout c was selected for further analysis due to its better (hyper-)parameter selection for the same method of the published plot.

@@

```{r}
#| echo: false
#| fig-cap: "XXX Number of bins for b is inadequate. Caption needs to start with what this figure is for XXX Model in \\gD{}, on the layout a and b, and two views of the fit in projections from $9\\text{-}D$, for the PBMC3k data ($n =  2622$ and $p = 9$). Layout a shows three-well separated clusters with big separation, while layout b shows close clusters. In $9\\text{-}D$, the data shows three clusters with two of them being close and one separated from the others. This is why b1, b2, and b3 shows a connected edge between the close clusters, as proven by @fig-pbmc-mse as a reasonable layout. Viewing the fitted model in $9\\text{-}D$, helps to see some unobserved patterns of the data: (i) dense points, and (ii) non-linear clusters. Videos of the langevitour animations are available at <https://youtu.be/0cKX_HG_n0k> and <https://youtu.be/KhJvsRtaX04> respectively."
#| label: fig-model-pbmc-author-proj
#| fig-pos: H
#| fig-width: 10
#| fig-height: 15
#| out-width: 80%

@@

<!--
XXX What this section needs:

- connect this with previous section, not make it seem like a completely new. connections are: nldr used to show clustering, scDEED provides a different approach to decide which is best.
- why these two? assume one is the published figure, and second is one recommended by scDEED, which our results confirm is better.
- Refer to figure in text
- Fig 16 not necessary

CHANGES WERE MADE BASED ON THIS THINKING
-->

<!-- In the field of single-cell studies, clustering is a common analytical task used to identify groups of cells with similar expression profiles. Non-linear dimensional reduction (NLDR) methods are frequently employed to visualize these clusters and help validate the results. However, it is well known that the 2D embeddings produced by t-SNE and UMAP may not accurately reflect the similarities among cell clusters. ALREADY STATED IN PREVIOUS SECTION -->

As we were writing this paper @xia2023 appeared proposing a new method called scDEED helping to assess the validity of a \gD{} embedding. scDEED calculates a reliability score for each cell embedding based on the similarity between the cell’s \gD{} embedding neighbors and its neighbors prior to embedding. A low reliability score suggests a dubious embedding. It can help in the deciding on optimal hyperparameters. Here we illustrate how our method compares with the results from scDEED

Following the process in @xia2023 again using the Human Peripheral Blood Mononuclear Cells (PBMC) data. Note that @xia2023 uses a different subset from @sec-pbmc, which contains $31,021$ cells including cell type labels, and the gene expression levels were in the unit of log-transformed UMI count per $10,000$. They focused on three sequencing methods (inDrops, DropSeq, and SeqWell) and four common cell types Cytotoxic T cell, CD4+T cell, CD14+ Monocyte, and B cell.

@@

@fig-model-pbmc-author-proj-scdeed shows layout from the original published article along with that suggested by scDEED to be more accurate. The log MSE vs binwidth plot illustrates that our approach would suggest that scDEED is correct here, that layout b is more accurately reflecting the cluster structure in the data.

@@

```{r}
#| label: fig-pbmc-mse-umap
#| fig-pos: H
#| out-width: 100%
#| fig-cap: "Assessing which of the 2 UMAP layouts with different hyperparameter settings (n_neighbors: $30$, min_dist: $0.3$ (red), n_neighbors: $80$, min_dist: $0.5$ (blue)) on the PBMC data is the better representation using MSE for varying binwidth ($a_1$). Colour used for the lines and points in the left plot and in the scatterplots represents UMAP layout (a, b). Of the two, layout b is optimal across all binwidths making it the best choice."

@@

```{r}
#| echo: false
#| eval: false
#| fig-cap: "Model in \\gD{}, on the layout a, and two views of the fit in projections from $50\\text{-}D$, for the PBMC data ($n =  5858$ and $p = 50$). Layout shows close clusters. In $50\\text{-}D$, the data shows "
#| label: fig-model-pbmc-author-proj-scdeed
#| fig-pos: H
#| fig-width: 15
#| fig-height: 5
#| out-width: 80%

### 2025-01-12 paper.tex

\includegraphics[width=0.6\textwidth,height=\textheight]{paper_files/figure-pdf/fig-p-d-error-in-2d-two-curvy-1.pdf}

@@

\caption{\label{fig-rm-lg}Distribution of edge lengths and wireframes in
\gD{} and \(7\text{-}D\) with Delaunay triangulation, and two choices of
long edge removal: (b) benchmark = 2.5\(a_1\) and (c) benchmark =
2\(a_1\), where \(a_1\) = 0.09.}

@@

\subsection{Single-cell gene
expression}\label{single-cell-gene-expression}

In the field of single-cell studies, a common analytical task involves
clustering to identify groups of cells with similar expression profiles.
NLDR methods are commonly used to display clusters, and help to verify
the results. For example, \citet{chen2023} illustrates the use of UMAP
to identify clusters in Human Peripheral Blood Mononuclear Cells
(PBMC3k). There are \(2622\) single cells. First 9 principal components
are used to generate the UMAP. Figure~\ref{fig-NLDR-variety} (a) is the
reproduction of the published plot. The objective is to assess the
published layout, and if it does not accurately represent the three
clusters with small separations of the PBMC3k dataset
(Figure~\ref{fig-model-pbmc-author-proj} (a2)), then select a reasonable
\gD{} layout.

@@

As a result, it is necessary to find an appropriate layout for the
dataset. MSE for different binwidths (\(a_1\)) using tSNE, UMAP, PHATE,
PaCMAP, and TriMAP with various (hyper-)parameter settings were computed
(Figure~\ref{fig-pbmc-mse}). Layouts c, d, and e, which show small
separations between clusters, are universally optimal. However, layout d
performs well with smaller binwidths and poorly with larger binwidths.
On the other hand, layout e performs well with larger binwidths. Layout
c was selected for further analysis due to its better (hyper-)parameter
selection for the same method of the published plot.

@@

\caption{\label{fig-model-pbmc-author-proj}Model in \gD{}, on the layout
a and b, and two views of the fit in projections from \(9\text{-}D\),
for the PBMC3k data (\(n =  2622\) and \(p = 9\)). Layout a shows
three-well separated clusters with big separation, while layout b shows
close clusters. In \(9\text{-}D\), the data shows three clusters with
two of them being close and one separated from the others. This is why
b1, b2, and b3 shows a connected edge between the close clusters, as
proven by Figure~\ref{fig-pbmc-mse} as a reasonable layout. Viewing the
fitted model in \(9\text{-}D\), helps to see some unobserved patterns of
the data: (i) dense points, and (ii) non-linear clusters. Videos of the
langevitour animations are available at
\url{https://youtu.be/0cKX_HG_n0k} and
\url{https://youtu.be/KhJvsRtaX04} respectively.}

@@

In the field of single-cell studies, clustering is a common analytical
task used to identify groups of cells with similar expression profiles.
Non-linear dimensional reduction (NLDR) methods are frequently employed
to visualize these clusters and help validate the results. However, it
is well known that the 2D embeddings produced by t-SNE and UMAP may not
accurately reflect the similarities among cell clusters.

To address this challenge, \citet{xia2023} introduces a statistical
method called scDEED, which detects unreliable cell embeddings generated
by \gD{} embedding techniques. scDEED calculates a reliability score for
each cell embedding based on the similarity between the cell's \gD{}
embedding neighbors and its neighbors prior to embedding. By identifying
embeddings with low reliability scores as dubious and those with high
scores as trustworthy, scDEED minimizes the number of questionable cell
embeddings. This approach also provides clear guidance for optimizing
the hyperparameters of an embedding method.

To verify that the \gD{} cell embeddings became better aligned with the
cell types after scDEED's optimization, the Human Peripheral Blood
Mononuclear Cells (PBMC) dataset was used. It contained \(31,021\) cells
with cell type labels, and the gene expression levels were in the unit
of log-transformed UMI count per \(10,000\). They focused on three
sequencing methods (inDrops, DropSeq, and SeqWell) and four common cell
types Cytotoxic T cell, CD4+T cell, CD14+ Monocyte, and B cell.

@@

@@

\caption{\label{fig-pbmc-mse-umap}Assessing which of the 2 UMAP layouts
with different hyperparameter settings (n\_neighbors: \(30\), min\_dist:
\(0.3\) (red), n\_neighbors: \(80\), min\_dist: \(0.5\) (blue)) on the
PBMC data is the better representation using MSE for varying binwidth
(\(a_1\)). Colour used for the lines and points in the left plot and in
the scatterplots represents UMAP layout (a, b). Layout b is universally
suboptimal. Layout b is the best choice.}

\includegraphics[width=0.8\textwidth,height=\textheight]{paper_files/figure-pdf/fig-model-pbmc-author-proj-scdeed-1.pdf}

\caption{\label{fig-model-pbmc-author-proj-scdeed}Model in \gD{}, on the
layout a, and two views of the fit in projections from \(50\text{-}D\),
for the PBMC data (\(n =  5858\) and \(p = 50\)). Layout shows close
clusters. In \(50\text{-}D\), the data shows}

>>>>

\includegraphics[width=0.6\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-p-d-error-in-2d-two-curvy-1.pdf}

@@

\caption{\label{fig-rm-lg}XXX Shouldn't the middle one have an edge
connecting inside a C, so that it is clearly not as good as c? XXX
Distribution of edge lengths and wireframes in \gD{} and \(7\text{-}D\)
with Delaunay triangulation, and two choices of long edge removal: (b)
benchmark = 2.5\(a_1\) and (c) benchmark = 2\(a_1\), where \(a_1\) =
0.09.}

@@

XXX WHY THESE EXAMPLES. NEEDS AN OVERVIEW PARAGRAPH

\subsection{Single-cell gene expression}\label{sec-pbmc}

XXX What is single cell data????

Clustering of single-cell data is used to identify groups of cells with
similar expression profiles. NLDR often used to summarise the discovered
clusters, and help to understand the results. The purpose of this
example is to \emph{illustrate how to use our method to help decide on
an appropriate NLDR layout that accurately represents the data}. The
cluster results of Human Peripheral Blood Mononuclear Cells (PBMC3k) in
\citet{chen2023} are examined. There are \(2622\) single cells, with XXX
gene expressions. Following their pre-processing, UMAP was performed
using nine principal components. Figure~\ref{fig-NLDR-variety} (a) is
the reproduction of the published plot, with colour indicating the
clustering results overlaid on the NLDR layout. The question is whether
this accurately represents the cluster structure in the data. Our method
can help here, and also help to provide a better \gD{} layout, as
needed.

@@

As a result, it is necessary to find an appropriate layout for the
dataset. MSE for different binwidths (\(a_1\)) using tSNE, UMAP, PHATE,
PaCMAP, and TriMAP with various (hyper-)\textasciitilde parameter
settings were computed (Figure~\ref{fig-pbmc-mse}). Layouts c, d, and e,
which show small separations between clusters, are universally optimal.
However, layout d performs well with smaller binwidths and poorly with
larger binwidths. On the other hand, layout e performs well with larger
binwidths. Layout c was selected for further analysis due to its better
(hyper-)parameter selection for the same method of the published plot.

@@

\caption{\label{fig-model-pbmc-author-proj}XXX Number of bins for b is
inadequate. Caption needs to start with what this figure is for XXX
Model in \gD{}, on the layout a and b, and two views of the fit in
projections from \(9\text{-}D\), for the PBMC3k data (\(n =  2622\) and
\(p = 9\)). Layout a shows three-well separated clusters with big
separation, while layout b shows close clusters. In \(9\text{-}D\), the
data shows three clusters with two of them being close and one separated
from the others. This is why b1, b2, and b3 shows a connected edge
between the close clusters, as proven by Figure~\ref{fig-pbmc-mse} as a
reasonable layout. Viewing the fitted model in \(9\text{-}D\), helps to
see some unobserved patterns of the data: (i) dense points, and (ii)
non-linear clusters. Videos of the langevitour animations are available
at \url{https://youtu.be/0cKX_HG_n0k} and
\url{https://youtu.be/KhJvsRtaX04} respectively.}

@@

As we were writing this paper \citet{xia2023} appeared proposing a new
method called scDEED helping to assess the validity of a \gD{}
embedding. scDEED calculates a reliability score for each cell embedding
based on the similarity between the cell's \gD{} embedding neighbors and
its neighbors prior to embedding. A low reliability score suggests a
dubious embedding. It can help in the deciding on optimal
hyperparameters. Here we illustrate how our method compares with the
results from scDEED

Following the process in \citet{xia2023} again using the Human
Peripheral Blood Mononuclear Cells (PBMC) data. Note that
\citet{xia2023} uses a different subset from Section~\ref{sec-pbmc},
which contains \(31,021\) cells including cell type labels, and the gene
expression levels were in the unit of log-transformed UMI count per
\(10,000\). They focused on three sequencing methods (inDrops, DropSeq,
and SeqWell) and four common cell types Cytotoxic T cell, CD4+T cell,
CD14+ Monocyte, and B cell.

@@

\textbf{?@fig-model-pbmc-author-proj-scdeed} shows layout from the
original published article along with that suggested by scDEED to be
more accurate. The log MSE vs binwidth plot illustrates that our
approach would suggest that scDEED is correct here, that layout b is
more accurately reflecting the cluster structure in the data.

@@

\caption{\label{fig-pbmc-mse-umap}Assessing which of the 2 UMAP layouts
with different hyperparameter settings (n\_neighbors: \(30\), min\_dist:
\(0.3\) (red), n\_neighbors: \(80\), min\_dist: \(0.5\) (blue)) on the
PBMC data is the better representation using MSE for varying binwidth
(\(a_1\)). Colour used for the lines and points in the left plot and in
the scatterplots represents UMAP layout (a, b). Of the two, layout b is
optimal across all binwidths making it the best choice.}
